---
title: "PSTAT 120B: Mathematical Statistics"
subtitle: "Ethan's Week 6 Discussion Section"
footer: "PSTAT 120B - Fall 2025 with Dr. Uma Ravat; material © Ethan P. Marzban"
logo: "Images/120b_logo.svg"
format: 
  clean-revealjs:
    theme: ../slides.scss
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
    menu:
      side: left
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: Ethan P. Marzban
    affiliations: Department of Statistics and Applied Probability; UCSB <br /> <br />
institute: November 7, 2025
title-slide-attributes:
    data-background-image: "Images/120b_logo.svg"
    data-background-size: "30%"
    data-background-opacity: "0.5"
    data-background-position: 80% 50%
code-annotations: hover
---

<style>
mjx-math {
  font-size: 80% !important;
}
</style>

<script>
MathJax = {
  options: {
    menuOptions: {
      settings: {
        assistiveMml: false
      }
    }
  }
};
</script>
<script type="text/javascript" id="MathJax-script" async src="path-to-MathJax/tex-chtml.js"></script>


$$
\newcommand\R{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\1}{1\!\!1}
\newcommand{\comp}[1]{#1^{\complement}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\SD}{\mathrm{SD}}
\newcommand{\vect}[1]{\vec{\boldsymbol{#1}}}
\newcommand{\tvect}[1]{\vec{\boldsymbol{#1}}^{\mathsf{T}}}
\newcommand{\hvect}[1]{\widehat{\boldsymbol{#1}}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\tmat}[1]{\mathbf{#1}^{\mathsf{T}}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\probto}{\stackrel{\mathrm{p}}{\longrightarrow}}
\newcommand{\distto}{\stackrel{\mathrm{d}}{\longrightarrow}}
\DeclareMathOperator*{\argmin}{\mathrm{arg} \ \min}
\newcommand{\iid}{\stackrel{\mathrm{i.i.d.}}{\sim}}
$$

```{css echo = F}
.hscroll {
  height: 100%;
  max-height: 600px;
  max-width: 2000px;
  overflow: scroll;
}
```

```{r setup, echo = F}
library(tidyverse)
library(countdown)
library(fixest)
library(modelsummary) # Make sure you have >=v2.0.0
library(GGally)
library(ggokabeito)
library(reshape2)
library(pander)
library(gridExtra)
library(cowplot)
library(palmerpenguins)
library(plotly)
library(tidymodels)
```

## {{< fa message >}} Announcements

1) _Information about regrades_
   -  Additional Text


## {{< fa backward-fast >}} Recap
### Framework for Estimation

-   **Goal:** estimate a particular population parameter.
-   Three key terms:
    -   [**Estimand:**]{.alert} another word for the parameter we are trying to estimate.
    -   [**Estimator:**]{.alert} a statistic being used to estimate the estimand.
        -   Another way to think about this: a "rule" used to estimate the parameter.
    -   [**Estimate:**]{.alert} a particular _realization_ (i.e. _observed instance_) of an estimator.

## {{< fa sign-hanging >}} Estimation
### An Analogy

-   Here's an analogy from _OpenIntro_ Statistics (a very good introductory statistics textbook!)

-   Suppose we want to catch a fish in a murky lake, using a spear.
    -   The location of the fish is akin to the _estimand_, and our spear throwing abilities are akin our _estimator_.
    
-   One strategy would be to throw the spear exactly where we last saw the fish.
    -   This is like constructing an _estimate_ from a realized dataset.

-   But, perhaps a _better_ strategy is to throw a _net_ in the vicinity of where we last saw the fish.
    -   So, what's the statistical analog of a net?
    
    
## {{< fa sign-hanging >}} Interval Estimation
### General Setup

-   Instead of reporting a _single_ value as our estimate, we report an _interval_ as our estimate.

::: {.fragment}
::: {.callout-note}
## **Definition**

An [**interval estimator**]{.alert} (aka [**confidence interval**]{.alert}) is an interval 
$$ \left[ \widehat{\theta}_{L} \ , \ \widehat{\theta}_{U} \right] $$
with _random_ endpoints (i.e. point estimators) $\widehat{\theta}_{L}$ and $\widehat{\theta}_{U}$.
:::
:::

-   We set up our intervals to have specified [**coverage probability**]{.alert}: $$\Prob( \widehat{\theta}_{L} \leq \theta \leq \widehat{\theta}_U) = 1 - \alpha $$


## {{< fa sign-hanging >}} Interval Estimation
### Two Results

-   [**Normal CI**]{.alert}: $\overline{Y}_n \pm \Phi^{-1}\left(1 - \frac{\alpha}{2} \right) \cdot \frac{\sigma}{\sqrt{n}}$
    -   Use if the population is either normally distributed, or the sample size is large enough for the CLT to kick in
    -   Use if the population standard deviation $\sigma$ is known.
    
-   [**_t_ CI**]{.alert}: $\overline{Y}_n \pm F_{t_{n - 1}}^{-1} \left(1 - \frac{\alpha}{2} \right) \cdot \frac{s}{\sqrt{n}}$
    -   Use if the population is either normally distributed, or the sample size is large enough for the CLT to kick in
    -   Use if the population standard deviation $\sigma$ is unknown.


## {{< fa sign-hanging >}} Interval Estimation
### Interpretation of Results

-   Say $[a, b]$ is a numerical 95\% confidence interval for $\theta$.

-   The interpretation of this interval is: we are 95\% confident that the interval $[a, b]$ covers the true value of $\theta$.
    -   It is **INCORRECT** to say that "there is a 95\% probability that $\theta$ is in the interval $[a, b]$," because none of these quantities ($a$, $b$, $\theta$) are random!
    
-   An equivalent interpretation: the true value of $\theta$ is estimated to be between $a$ and $b$ for 95\% of all intervals constructed from all possible samples.


## {{< fa sign-hanging >}} Interval Estimation
### Interpretation of Results

```{r}
set.seed(120)
lowers <- c()
uppers <- c()
centers <- c()
for(b in 1:100){
  temp_samp <- rnorm(25, 2, 1.5)
  centers <- c(centers, mean(temp_samp))
  lowers <- c(lowers, mean(temp_samp) - 1.96 * (1.5 / 5))
  uppers <- c(uppers, mean(temp_samp) + 1.96 * (1.5 / 5))
}

df <- data.frame(
  lowers,
  centers,
  uppers,
  x = 1:length(centers)
) %>%
  mutate(covind = ifelse((lowers >= 2) | (uppers <= 2), "no", "yes"))


df %>%
  ggplot(aes(x = x)) +
  geom_segment(aes(x = x, xend = x, y = lowers, yend = uppers,
                   col = covind),
               linewidth = 0.75) +
  theme_minimal(base_size = 18) +
  scale_color_manual(values = c("red", "blue")) + 
  ggtitle("100 CIs for a Mean",
          subtitle = "Normal(25, 2.25) Population; n = 25") +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold",
                             size = 24))  +
  xlab("") + ylab("")
```



## {{< fa sign-hanging >}} Interval Estimation
### Interpretation of Results

```{r}
set.seed(130)
lowers <- c()
uppers <- c()
centers <- c()
for(b in 1:100){
  temp_samp <- rnorm(25, 2, 1.5)
  centers <- c(centers, mean(temp_samp))
  lowers <- c(lowers, mean(temp_samp) - 1.96 * (1.5 / 5))
  uppers <- c(uppers, mean(temp_samp) + 1.96 * (1.5 / 5))
}

df <- data.frame(
  lowers,
  centers,
  uppers,
  x = 1:length(centers)
) %>%
  mutate(covind = ifelse((lowers >= 2) | (uppers <= 2), "no", "yes"))


df %>%
  ggplot(aes(x = x)) +
  geom_segment(aes(x = x, xend = x, y = lowers, yend = uppers,
                   col = covind),
               linewidth = 0.75) +
  theme_minimal(base_size = 18) +
  scale_color_manual(values = c("red", "blue")) + 
  ggtitle("100 CIs for a Mean",
          subtitle = "Normal(25, 2.25) Population; n = 25") +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold",
                             size = 24))  +
  xlab("") + ylab("")
```


## {{< fa sign-hanging >}} Interval Estimation
### Interpretation of Results

```{r}
set.seed(100)
lowers <- c()
uppers <- c()
centers <- c()
for(b in 1:100){
  temp_samp <- rnorm(25, 2, 1.5)
  centers <- c(centers, mean(temp_samp))
  lowers <- c(lowers, mean(temp_samp) - 1.96 * (1.5 / 5))
  uppers <- c(uppers, mean(temp_samp) + 1.96 * (1.5 / 5))
}

df <- data.frame(
  lowers,
  centers,
  uppers,
  x = 1:length(centers)
) %>%
  mutate(covind = ifelse((lowers >= 2) | (uppers <= 2), "no", "yes"))


df %>%
  ggplot(aes(x = x)) +
  geom_segment(aes(x = x, xend = x, y = lowers, yend = uppers,
                   col = covind),
               linewidth = 0.75) +
  theme_minimal(base_size = 18) +
  scale_color_manual(values = c("red", "blue")) + 
  ggtitle("100 CIs for a Mean",
          subtitle = "Normal(25, 2.25) Population; n = 25") +
  theme(legend.position = "none",
        plot.title = element_text(face = "bold",
                             size = 24))  +
  xlab("") + ylab("")
```




## {{< fa arrow-right >}} Consistency
### Definition

::: {.callout-important}
## **Definition:** Consistency

An estimator $\widehat{\theta}_n$ is said to be a [**consistent**]{.alert} estimator for $\theta$, denoted
$$ \widehat{\theta}_n \probto \theta $$
if, for any $\varepsilon > 0$, either of the following equivalent statements hold:
\begin{align*}
  \lim_{n \to \infty} \Prob\left( |\widehat{\theta}_n - \theta| \leq \varepsilon \right)  & = 1 \\
  \lim_{n \to \infty} \Prob\left( |\widehat{\theta}_n - \theta| >  \varepsilon \right)  & = 0
\end{align*}
:::

-   Let's break this down.


## {{< fa arrow-right >}} Consistency
### Interpretation

-   $|\widehat{\theta}_n - \theta|$ is essentially the _distance_ between $\widehat{\theta}_n$ and $\theta$.

-   Saying that $|\widehat{\theta}_n - \theta| \leq \varepsilon$ for any $\varepsilon > 0$ is saying that "the distance between $\widehat{\theta}_n$ and $\theta$ is very small."
    -   Equivalently: "$\widehat{\theta}_n$ is very close to $\theta$."
    
-   The definition of consistency asserts that this probability goes to zero as the sample size increases.
    -   That is: "as our sample size becomes larger, we become more certain that $\widehat{\theta}_n$ is very close to $\theta$."
    
-   Seems like a pretty nice property for an estimator to have, doesn't it?
    
    
## {{< fa arrow-right >}} Consistency
### Consistency and Unbiasedness

-   **Question:** are all consistent estimators unbiased?
    -   **Answer:** _no_! For example, $S_n^2 := \frac{1}{n} \sum_{i=1}^{n} (Y_i - \overline{Y}_n)^2$ is a biased estimator for $\sigma^2$, but it turns out it is a consistent estimator for $\sigma^2$.
    
-   **Question:** are all unbiased estimators consistent?
    -   **Answer:** you tell me...
    
    
::: {.fragment}
::: {.callout-important}
## **Theorem 9.1**
An unbiased estimator $\widehat{\theta}_n$ for $\theta$ is consistent if $\lim\limits_{n \to \infty} \Var(\widehat{\theta}_n) = 0$.
:::
:::
