---
title: "PSTAT 120B: Mathematical Statistics"
subtitle: "Ethan's Week 7 Discussion Section"
footer: "PSTAT 120B - Fall 2025 with Dr. Uma Ravat; material © Ethan P. Marzban"
logo: "Images/120b_logo.svg"
format: 
  clean-revealjs:
    theme: ../slides.scss
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
    menu:
      side: left
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: Ethan P. Marzban
    affiliations: Department of Statistics and Applied Probability; UCSB <br /> <br />
institute: November 14, 2025
title-slide-attributes:
    data-background-image: "Images/120b_logo.svg"
    data-background-size: "30%"
    data-background-opacity: "0.5"
    data-background-position: 80% 50%
code-annotations: hover
---

<style>
mjx-math {
  font-size: 80% !important;
}
</style>

<script>
MathJax = {
  options: {
    menuOptions: {
      settings: {
        assistiveMml: false
      }
    }
  }
};
</script>
<script type="text/javascript" id="MathJax-script" async src="path-to-MathJax/tex-chtml.js"></script>


$$
\newcommand\R{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\1}{1\!\!1}
\newcommand{\comp}[1]{#1^{\complement}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\SD}{\mathrm{SD}}
\newcommand{\vect}[1]{\vec{\boldsymbol{#1}}}
\newcommand{\tvect}[1]{\vec{\boldsymbol{#1}}^{\mathsf{T}}}
\newcommand{\hvect}[1]{\widehat{\boldsymbol{#1}}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\tmat}[1]{\mathbf{#1}^{\mathsf{T}}}
\newcommand{\Cov}{\mathrm{Cov}}
\newcommand{\probto}{\stackrel{\mathrm{p}}{\longrightarrow}}
\newcommand{\distto}{\stackrel{\mathrm{d}}{\longrightarrow}}
\DeclareMathOperator*{\argmin}{\mathrm{arg} \ \min}
\newcommand{\iid}{\stackrel{\mathrm{i.i.d.}}{\sim}}
$$

```{css echo = F}
.hscroll {
  height: 100%;
  max-height: 600px;
  max-width: 2000px;
  overflow: scroll;
}
```

```{r setup, echo = F}
library(tidyverse)
library(countdown)
library(fixest)
library(modelsummary) # Make sure you have >=v2.0.0
library(GGally)
library(ggokabeito)
library(reshape2)
library(pander)
library(gridExtra)
library(cowplot)
library(palmerpenguins)
library(plotly)
library(tidymodels)
```


## {{< fa arrow-right >}} Consistency
### Definition

::: {.callout-note}
## **Definition:** Consistency

An estimator $\widehat{\theta}_n$ for $\theta$ is said to be a [**consistent estimator**]{.alert} for $\theta$ if either of the following equivalent conditions hold: for any $\varepsilon > 0$,
\begin{align*}
  \lim_{n \to \infty} \Prob(|\widehat{\theta}_n - \theta| \leq \varepsilon)  & = 1 \\
  \lim_{n \to \infty} \Prob(|\widehat{\theta}_n - \theta| > \varepsilon)  & = 0
\end{align*}
:::

-   This definition can seem scary at first; let's break it down.

## {{< fa arrow-right >}} Consistency
### Definition: Interpretation

-   $|\widehat{\theta}_n - \theta|$ is essentially the distance between our estimator and estimand.

-   The event $\{|\widehat{\theta}_n - \theta| < \varepsilon\}$ is then: "the distance between our estimator and estimand is small."
    -   In other words: "our estimator is very close to our estimand."
    
-   Saying that $\Prob(|\widehat{\theta}_n - \theta| < \varepsilon) \to 1$ as $n \to \infty$ means that, as our sample size gets large, we are more and more certain that our estimator is very close to our estimand.

-   That's all consistency is saying!
    -   I find this breakdown useful when trying to remember the definition of consistency.

## {{< fa arrow-right >}} Consistency
### Main Theorem

-   We don't always need to use the definition of consistency:

::: {.fragment}
::: {.callout-important}
## **Theorem**

If $\widehat{\theta}_n$ is an unbiased estimator for $\theta$, then it is also a consistent estimator for $\theta$ if and only if $\Var(\widehat{\theta}_n) \to 0$ as $n \to \infty$.
:::
:::

-   **Caution:** If you want to use this theorem, you MUST first check that your estimator is unbiased! (Otherwise, we can't appeal to this theorem.)

-   **Question:** are all unbiased estimators consistent?

-   **Question:** are all consistent estimators unbiased?


## {{< fa arrow-right >}} Consistency
### Biased but Consistent

**Example:** $\widehat{\sigma^2}_n := \frac{1}{n} \sum_{i=1}^{n} (Y_i - \bar{Y}_n)^2$
```{r}
exact_dist <- \(x, n, sig2){
  dgamma(x, shape = (n - 1)/2, scale = 2*sig2/n)
}

data.frame(x = 0:6) %>% ggplot(aes(x = x)) +
  stat_function(fun = exact_dist,
                args = list(n = 5,
                            sig2 = 1.5),
                aes(colour = "05"),
                linewidth = 1.5
  ) + stat_function(fun = exact_dist,
                args = list(n = 10,
                            sig2 = 1.5),
                aes(colour = "10"),
                linewidth = 1.5
  ) + stat_function(fun = exact_dist,
                args = list(n = 15,
                            sig2 = 1.5),
                aes(colour = "15"),
                linewidth = 1.5
  ) + stat_function(fun = exact_dist,
                args = list(n = 20,
                            sig2 = 1.5),
                aes(colour = "20"),
                n = 200,
                linewidth = 1.5
  ) + stat_function(fun = exact_dist,
                args = list(n = 25,
                            sig2 = 1.5),
                aes(colour = "25"),
                n = 200,
                linewidth = 1.5
  )  +
  geom_vline(xintercept = 1.5, linetype = "dashed") +
  theme_minimal(base_size = 18) +
  ggtitle("Modified Sample Variance") +
  labs(colour = "n") +
  theme(plot.title = element_text(face = "bold",
                                  size = 24)) +
  scale_color_okabe_ito()
```

## {{< fa dice >}} Likelihoods

-   A coin lands "heads" with probability $\theta \in (0, 1)$. 

-   Suppose we toss the coin five times, and observe (`H`, `T`, `H`, `H`, `T`).
    -   Letting _X_ denote the number of heads in 5 tosses, _X_ ~ Bin(5, $\theta$)

-   If $\theta = 0.3$, how likely were we to have observed what we saw?
    -   If $\theta = 0.3$, the probability of observing three heads is $\Prob(X = 3) = \binom{5}{3}(0.3)^3(0.7)^2 \approx 0.1323$
    
-   If $\theta = 0.4$, how likely were we to have observed what we saw?
    -   If $\theta = 0.3$, the probability of observing three heads is $\Prob(X = 3) = \binom{5}{3}(0.4)^3(0.6)^2 \approx 0.2304$
    
-   Given that we observed (`H`, `T`, `H`, `H`, `T`), which was more likely: that $\theta = 0.3$ or that $\theta = 0.4$?

## {{< fa dice >}} Likelihoods
### Maximization

::: {.callout-tip}
## **Idea:** 

To estimate a parameter $\theta$ given data $Y_1, Y_2, \cdots, Y_n$, find the value of $\theta$ that was _most likely_ to have given rise to the data we observed.
:::

-   This procedure is called [**maximum likelihood estimation**]{.alert} and the resulting estimator is called the [**maximum likelihood estimator**]{.alert}

-   The [**likelihood**]{.alert} gives a way of answering the question: "given data $Y_1, Y_2, \cdots, Y_n$, how likely was it that the true value of the parameter were $\theta$, for an arbitrary $\theta$?"

-   For example, back to our coin-tossing example:


## {{< fa dice >}} Likelihoods
### Example

```{r}
#| echo: False
#| code-fold: True

lik_theta <- Vectorize(function(theta){
  dbinom(3, 5, theta)
})

data.frame(x = 0:1) %>% ggplot(aes(x = x)) +
  stat_function(fun = lik_theta, linewidth = 1) +
  theme_minimal(base_size = 18) +
  xlab(bquote(theta)) +
  ylab(paste("likelihood that true prob. is", bquote(theta))) +
  ggtitle("Example Likelihood",
          subtitle = "Data = (H, T, H, H, T)") +
  theme(plot.title = element_text(face = "bold",
                                  size = 24)) +
  annotate(
    "point",
    x = 0.3, y = lik_theta(0.3),
    size = 4
  ) + annotate(
    "point",
    x = 0.4, y = lik_theta(0.4),
    size = 4
  ) +
  geom_segment(
    aes(x = 0.3, xend = 0.3,
        y = 0, yend = lik_theta(0.3)),
    linetype = "dotted"
  ) + geom_segment(
    aes(x = 0, xend = 0.3,
        y = lik_theta(0.3), yend = lik_theta(0.3)),
    linetype = "dotted"
  ) +
  geom_segment(
    aes(x = 0.4, xend = 0.4,
        y = 0, yend = lik_theta(0.4)),
    linetype = "dotted"
  ) + geom_segment(
    aes(x = 0, xend = 0.4,
        y = lik_theta(0.4), yend = lik_theta(0.4)),
    linetype = "dotted"
  )
```


## {{< fa dice >}} Likelihoods
### Example: Different Datasets

```{r}
#| echo: False
#| code-fold: True

lik_theta <- function(theta, x){
  dbinom(sum(x == "H"), length(x), theta)
}

data.frame(x = 0:1) %>% ggplot(aes(x = x)) +
  stat_function(fun = lik_theta, linewidth = 1.5,
                args = list(x = c("H", "T", "H", "H", "T")),
                aes(colour = "c(H, T, H, H, T)")
  ) + stat_function(fun = lik_theta, linewidth = 1.5,
                args = list(x = c("T", "T", "T", "T", "H")),
                aes(colour = "c(T, T, T, T, H)")
  ) + stat_function(fun = lik_theta, linewidth = 1.5,
                args = list(x = c("H", "H", "H", "H", "T")),
                aes(colour = "c(H, H, H, H, T)")
  ) +
  theme_minimal(base_size = 18) +
  labs(colour = "Data") +
  ggtitle("Example Likelihood",
          subtitle = "Different Datasets") +
  theme(plot.title = element_text(face = "bold",
                                  size = 24)) +
  xlab(bquote(theta)) +
  ylab(paste("likelihood that true prob. is", bquote(theta))) +
  scale_color_okabe_ito()
```


## {{< fa dice >}} Likelihoods
### Definition

-   The likelihood given a sample is just the marginal mass/density function.

-   Indeed, the reason we call this a "likelihood" function instead of a "probability" function is because, in the continuous case, the likelihood at a given point doesn't really represent a true _probability_ (but, it is giving the same _sort_ of information as a probability).

::: {.fragment}
$$ \mathcal{L}(\theta; Y_1, Y_2, \cdots, Y_n) := f_{Y_1, Y_2, \cdots, Y_n}(Y_1, Y_2, \cdots, Y_n; \theta) $$
:::

-   If the sample is i.i.d.: $\displaystyle \mathcal{L}(\theta; Y_1, Y_2, \cdots, Y_n) = \prod_{i=1}^{n} f(Y_i; \theta)$

## {{< fa dice >}} Likelihoods
### Continuous Example: Exponential

-   **Example Scenario:** The wait time at _The Arbor_ of a randomly-selected customer follows an Exponential distribution with unknown mean $\theta$. 

-   Likelihood (assuming i.i.d.):
\begin{align*}
  \class{fragment}{{} \mathcal{L}(\theta; Y_1, Y_2, \cdots, Y_n)}
    &\class{fragment}{{} = \prod_{i=1}^{n} f(Y_i; \theta)}            \\[3px]
    &\class{fragment}{{} = \prod_{i=1}^{n} \left[ \frac{1}{\theta} e^{-Y_i/\theta} \cdot 1 \! \! 1_{\{Y_i \geq 0\}} \right] }    \\[3px]
    &\class{fragment}{{} = \left( \frac{1}{\theta} \right)^n \cdot e^{-\frac{1}{\theta} \sum_{i=1}^{n} Y_i} \cdot 1 \! \! 1_{\{Y_{(1)} \geq 0\}}} 
\end{align*}

## {{< fa dice >}} Likelihoods
### Continuous Example: Exponential

$$ \mathcal{L}(\theta; Y_1, Y_2, \cdots, Y_n) = \left( \frac{1}{\theta} \right)^n \cdot e^{-\frac{1}{\theta} \sum_{i=1}^{n} Y_i} \cdot 1 \! \! 1_{\{Y_{(1)} \geq 0\}} $$

-   Note that we are viewing this as a function of $\theta$! 
    -   The idea is: we assume that we already have our data $(Y_1, Y_2, \cdots, Y_n)$; what we're trying to do is figure out the _most plausible_ $\theta$ value given the data we've observed.
    
-   Further note that this means likelihoods aren't necessarily true density functions in $\theta$.

## {{< fa dice >}} Likelihoods
### Continuous Example: Exponential

```{r}
#| echo: True
#| code-fold: True
exp_lik <- function(theta, y) {
  (1 / theta)^(length(y)) * exp(-(1 / theta) * sum(y))
}
```

```{r}
data.frame(x = 0:6) %>% ggplot(aes(x = x)) +
  stat_function(fun = exp_lik,
                args = list(y = c(1, 2, 3, 1, 2)),
                aes(colour = "c(1, 2, 3, 1, 2)"),
                linewidth = 1.5
  ) + stat_function(fun = exp_lik,
                args = list(y = c(2, 2, 2, 2, 2)),
                aes(colour = "c(2, 2, 2, 2, 2)"),
                linewidth = 1.5
  ) + stat_function(fun = exp_lik,
                args = list(y = c(1, 2, 3, 2, 2)),
                aes(colour = "c(1, 2, 3, 2, 2)"),
                linewidth = 1.5,
                linetype = "dashed"
  ) +
  theme_minimal(base_size = 18) +
  labs(colour = "Data") +
  ggtitle("Exponential Likelihood",
          subtitle = "Different Datasets") +
  theme(plot.title = element_text(face = "bold",
                                  size = 24)) +
  xlab(bquote(theta)) +
  ylab(paste("likelihood that true avg. is", bquote(theta))) +
  scale_color_manual(values = c("#009E73","#0072B2","#E69F00")) 
```

