---
title: "PSTAT 120B: Mathematical Statistics"
subtitle: "Ethan's Week 3 Discussion Section"
footer: "PSTAT 120B - Fall 2025 with Dr. Uma Ravat; material © Ethan P. Marzban"
logo: "Images/120b_logo.svg"
format: 
  clean-revealjs:
    theme: ../slides.scss
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
    menu:
      side: left
html-math-method:
  method: mathjax
  url: "https://cdn.jsdelivr.net/npm/mathjax@3/es5/tex-mml-chtml.js"
author:
  - name: Ethan P. Marzban
    affiliations: Department of Statistics and Applied Probability; UCSB <br /> <br />
institute: October 17, 2025
title-slide-attributes:
    data-background-image: "Images/120b_logo.svg"
    data-background-size: "30%"
    data-background-opacity: "0.5"
    data-background-position: 80% 50%
code-annotations: hover
---

<style>
mjx-math {
  font-size: 80% !important;
}
</style>

<script>
MathJax = {
  options: {
    menuOptions: {
      settings: {
        assistiveMml: false
      }
    }
  }
};
</script>
<script type="text/javascript" id="MathJax-script" async src="path-to-MathJax/tex-chtml.js"></script>


$$
\newcommand\R{\mathbb{R}}
\newcommand{\N}{\mathbb{N}}
\newcommand{\E}{\mathbb{E}}
\newcommand{\Prob}{\mathbb{P}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\1}{1\!\!1}
\newcommand{\comp}[1]{#1^{\complement}}
\newcommand{\Var}{\mathrm{Var}}
\newcommand{\SD}{\mathrm{SD}}
\newcommand{\vect}[1]{\vec{\boldsymbol{#1}}}
\newcommand{\tvect}[1]{\vec{\boldsymbol{#1}}^{\mathsf{T}}}
\newcommand{\hvect}[1]{\widehat{\boldsymbol{#1}}}
\newcommand{\mat}[1]{\mathbf{#1}}
\newcommand{\tmat}[1]{\mathbf{#1}^{\mathsf{T}}}
\newcommand{\Cov}{\mathrm{Cov}}
\DeclareMathOperator*{\argmin}{\mathrm{arg} \ \min}
\newcommand{\iid}{\stackrel{\mathrm{i.i.d.}}{\sim}}
$$

```{css echo = F}
.hscroll {
  height: 100%;
  max-height: 600px;
  max-width: 2000px;
  overflow: scroll;
}
```

```{r setup, echo = F}
library(tidyverse)
library(countdown)
library(fixest)
library(modelsummary) # Make sure you have >=v2.0.0
library(GGally)
library(ggokabeito)
library(reshape2)
library(pander)
library(gridExtra)
library(cowplot)
library(palmerpenguins)
library(plotly)
library(tidymodels)
```

## {{< fa chart-simple >}} Sampling Distributions
### Overview

-   Given a sample _Y_~1~, ..., _Y_~_n_~, a [**statistic**]{.alert} is a function _h_(_Y_~1~, ..., _Y_~_n_~) of the sample.
    -   Statistics are random, and therefore have distributions; a statistics' distribution is called its [**sampling distribution**]{.alert}
    
-   How do we find sampling distributions?
    -   Using our material from the Transformations chapter of this course!
    -   After all, statistics are just functions of random variables.
    
-   We start with the case of a normal population; i.e. that our sample is an i.i.d. sample drawn from a normal distribution.
    
## {{< fa chart-simple >}} Sampling Distributions
### Normal Population

-   Suppose $Y_1, \cdots, Y_n \iid \mathcal{N}(\mu, \sigma^2)$.

-   What is the sampling distribution of the sample mean
$$ \overline{Y}_n := \frac{1}{n} \sum_{i=1}^{n} Y_i $$ 

-   What do we know about the sampling distribution of the sample variance
$$ S_n^2 := \frac{1}{n - 1} \sum_{i=1}^{n} (Y_i - \overline{Y}_n)^2 $$


## {{< fa chart-simple >}} Sampling Distributions
### Normal Population: Sample Mean

```{r}
#| echo: True
#| code-fold: True
set.seed(120); means <- c(); n <- 25
for(k in 1:1000) {
  means <- c(means, mean(rnorm(n, 3, 1.2)))
}
```

```{r}
means %>% data.frame() %>% 
  ggplot(aes(x = `.`)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 15, col = "white") +
  theme_minimal(base_size = 24) + 
  xlab("x") +
  stat_function(fun = dnorm,
                args = list(mean = 3,
                            sd = sqrt(1.2 / n)),
                col = "blue", linewidth = 1.5) +
  ggtitle("Sample Mean", 
          subtitle = "Empirical Sampling Distribution") +
  theme(plot.title = element_text(face = "bold"))
```



## {{< fa chart-simple >}} Sampling Distributions
### Normal Population: Sample Variance

```{r}
#| echo: True
#| code-fold: True
set.seed(120); var_stat <- c(); n <- 25
for(k in 1:1000) {
  var_stat <- c(var_stat, ((n - 1) / (1.2^2)) * var(rnorm(n, 3, 1.2)))
}
```

```{r}
var_stat %>% data.frame() %>% 
  ggplot(aes(x = `.`)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 15, col = "white") +
  theme_minimal(base_size = 24) + 
  xlab("x") + 
  stat_function(fun = dchisq,
                args = list(df = n - 1),
                col = "blue", linewidth = 1.5) +
  ggtitle("Scaled Sample Variance", 
          subtitle = "Empirical Sampling Distribution") +
  theme(plot.title = element_text(face = "bold"))
```


## {{< fa bell >}} Standard Normal Distribution
### Quick Review

:::: {.columns}

::: {.column width="50%"}
-   $Z \sim \mathcal{N}(0, 1)$

-   **PDF:** $\displaystyle \phi(z) := \frac{1}{\sqrt{2\pi}} e^{-\frac{1}{2} z^2}$

-   **CDF:** $\displaystyle \Phi(z) := \int_{-\infty}^{z} \frac{1}{\sqrt{2\pi}} e^{-\frac{1}{2} x^2} \ \mathrm{d}x$

::: {.fragment}
```{r}
#| fig-height: 5

data.frame(x = -3:3) %>% 
  ggplot(aes(x = x)) +
  stat_function(fun = dnorm, 
                linewidth = 2,
                col = "blue") +
  theme_minimal(base_size = 28) +
  ggtitle("PDF") +
  xlab("z") + ylab(bquote(phi(z)))
```
:::


:::

::: {.column width="50%"}
::: {.fragment}
```{r}
#| fig-height: 5

data.frame(x = -3:3) %>% 
  ggplot(aes(x = x)) +
  stat_function(fun = dnorm, 
                linewidth = 2,
                col = "blue") + 
  stat_function(geom = "area",
                fun = dnorm, 
                fill = "#E69F00",
                alpha = 0.8,
                xlim = c(-3, 1),
                linewidth = 2,
                col = "blue") +
  theme_minimal(base_size = 28) +
  ggtitle("CDF") +
  xlab("x") + ylab(bquote(phi(x))) 
```
:::


::: {.fragment}
```{r}
#| fig-height: 5

data.frame(x = -3:3) %>% 
  ggplot(aes(x = x)) +
  stat_function(fun = pnorm, 
                linewidth = 2,
                col = "blue") +
  theme_minimal(base_size = 28) +
  ggtitle("CDF") +
  xlab("z") + ylab(bquote(Phi(z))) 
```
:::
:::

::::

## {{< fa table >}} Lookup Tables
### Normal and Chi-Squared

-   $\Phi(z)$ has no closed-form expression; as such, it must be (in general) evaluated numerically. (One exception: $\Phi(0)$.)
    -   This means we either have to use a computer, or a **table**
    
-   Our table (for PSTAT 120B) gives **right-tail areas**, meaning it actually gives values of $1 - \Phi(z)$.
    -   This is likely the opposite of the tables you saw in PSTAT 120A.
    
-   **Example:** Evaluate $\Phi(0.53)$ using the table.

-   **Example:** Evaluate $\Phi^{-1}(0.9265)$ using the table.

-   **Example:** If $U \sim \chi^2_{8}$, find $c$ such that $\Prob(U > c) = 0.05$.